Extending statistical models for source attribution of zoonotic diseases: A study of campylobacteriosis

### 2018-03-20 paper/finalish/article_1023.tex
adds some acknowledgements
updates paper following review

\begin{document}
\runninghead{Extending statistical models for the source attribution of zoonotic diseases}
\title{Extending statistical models for source attribution of zoonotic diseases: A study of campylobacteriosis}
\author{Sih-Jing Liao\affil{a},
Jonathan Marshall\affil{a}\corrauth\ \footnotemark[2], Martin L. Hazelton\affil{a} and Nigel P. French\affil{b}}
\address{\affilnum{a} Institute of Fundamental Sciences-Statistics, Massey University, Palmerston North, New Zealand\\
\affilnum{b} mEpiLab, Hopkirk Research Institute, Massey University, Palmerston North, New Zealand}
\corraddr{Jonathan Marshall, Institute of Fundamental Sciences-Statistics, Massey University, Private Bag 11222, Palmerston North, New Zealand\\ \footnotemark[2] E-mail: J.C.marshall@massey.ac.nz}

@@

Preventing and controlling zoonoses through the design and implementation of public health policies requires a thorough understanding of transmission pathways. Modeling jointly the epidemiological data and genetic information on the microbial isolates derived from cases provides a methodology for tracing back the source of infection. The isolate genetic information and the epidemiological data may be modeled at a variety of levels. In this paper, the attribution probability for human cases of campylobacteriosis for each source, conditional on the extent to which each case resides in a rural compared to urban environment, is estimated. A model that incorporates genetic data and evolutionary processes is applied alongside a newly developed genetic data-free model. We show that the inference from each model is comparable except for rare microbial genotypes, and that the effect of ``rurality'' may be modeled linearly on the logit scale, with increasing rurality leading to increasing likelihood of ruminant-sourced campylobacteriosis.

@@

\section{Introduction}
Modeling of disease surveillance data to explore patterns of infectious diseases has had a long history in public health. Infectious diseases can cause high economic and medical costs due to morbidity and mortality. In recent decades, the annual number of global deaths caused by infections has leveled off at approximate 15 million and may remain at this level for the next three decades \cite{DyeC, WHOM}. In order for such an enormous health burden to be reduced, preventing and controlling infectious diseases becomes extraordinarily important, which depends on how much we know about the nature of disease transmission. For zoonotic diseases, where transmission to humans from animal reservoirs may be via food, water, through environmental contamination or via direct contact with animals, knowledge of the potential sources and pathways of infection is key to reducing the burden of disease. Quantifying the burden of illness attributable to each pathway and source provides valuable information to health professionals making decisions on public health policy for disease control.

Over the last few decades there have been many models developed for the analysis of zoonotic diseases. Epidemiological models make use of the incidence of cases and geographic data to capture the main characteristics of disease spread. For example, the Susceptible-Infectious-Recovered model \cite{Kerm} classifies populations into a susceptible, infective and recovered status enabling an estimation of the number of individuals who are likely to be infected. Also, models with a combination of spatial and temporal data enable the prediction of the time frame and areas where a disease outbreak is likely to occur \cite{HeldL, Hoehl, Simo}. However zoonotic transmission from animals to humans may be complex, involving many sources and exposures linked by differing pathways. For instance, infected wild birds may contaminate environmental water and cause disease spread to water users, either humans or other animals \cite{Wagen}. Tracing the source of infection is thus crucial to increasing the ability to implement risk management and intervention \cite{Wilso, Morel}. In terms of statistical analysis, modeling zoonoses requires an advanced approach with the focus changed from just epidemiology to a combination of epidemiology, evolutionary genetics and biology \cite{Muell}. Hence, some source attribution models have been proposed to estimate the number of cases attributable to different sources by utilizing epidemiological information and the association with genotypes found in humans and sources \cite{vanP, Hald, MullA}.

\begin{table}
  \begin{center}
    \begin{tabular}{crrrrrrr}
      \toprule
      ST & \textit{asp}A & \textit{gln}A & \textit{glt}A & \textit{gly}A & \textit{pgm}& \textit{tkt}& \textit{unc}A\\ \midrule
      403 & 10 & 27 & 16 & 19 & 10 & 5 & 7\\
      2026 & 10 & 1 & 16 & 19 & 10 & 5 & 7\\
      474 & 2 & 4 & 1 & 2 & 2 & 1 & 5 \\
      2343 & 2 & 4 & 5 & 2 & 10 & 1 & 5\\
      \bottomrule
    \end{tabular}
  \end{center}
  \caption{The allelic profile of genotypes ST-45 and ST-474 is composed of seven allele numbers at each of the seven loci.}
  \label{tab:mlst}
\end{table}

The genetic information used in such integrated models is typically derived from molecular genotyping that groups closely related organisms together \cite{Cotta}. Typically molecular typing methods, such as multilocus sequence typing (MLST) \cite{Dingl, Coll, Urwin} might be used, which utilizes nucleotide sequences of internal fragments of a small set of housekeeping genes. Such sequences have sufficient variation to distinguish differing pathogen lineages, while being relatively stable within lineages. Each unique nucleotide sequence (allele) at each housekeeping gene (locus) is assigned a number, and the set of numbers across all loci (the allelic profile) is then taken as the genotype, with is usually also assigned a number.

@@

In this paper we develop statistical models for the source attribution, and use a case study of human campylobacteriosis conducted in New Zealand \cite{Marsh} as a demonstration. Human campylobacteriosis is caused mainly by \textit{C. jejuni} and \textit{C. coli} which are the dominant species associated with 80\% and approximately 15\% of illnesses, respectively \cite{Guert}. The common symptoms of infection are diarrhea, abdominal pain and fever. However, a severe complication named Guillain Barre syndrome may develop, which is a life-threatening disease that weakens the nervous system and leads to paralysis in limbs and respiration \cite{Hanha}. The pathogen can be spread between animals, or from animals and wild birds to humans. The transmission route can be via drinking contaminated water, eating undercooked animal food products, or handling animal food products that are already contaminated by faeces. In this study, we compare the performance of the asymmetric Island model \cite{Wilso}, which considers genetic evolution when estimating the genotype sampling distribution on each source, to a simple model that utilizes only the prevalence of each type to derive the sampling distribution. We then extend both models to incorporate covariates such as rurality indicating the area where a human case resides, exploring the effect of human case rurality on attribution results, modeling attribution by rurality using a linear trend on the logit scale or using separate categories in a Bayesian context, and performing model comparison.

\section{Models for source attribution with and without microbial genetic information}
\subsection{Notation and assumptions}
The ultimate goal of attribution models is to estimate the probability that the observed human cases arise from each putative source. Our data comprise microbial genotype information from each observed human case obtained from analysis of stool samples, and also from a pool of non-human cases corresponding to potential zoonotic sources of disease

Given this genotyping information, we first estimate the sampling distribution of genotypes for each source, and then estimate the appropriate combinations of those genotype distributions that most likely give rise to the set of genotypes observed among human cases. Specifying first the sampling distribution of genotypes found on sources is fundamental for the purpose of not only exploring how it affects the source attribution probability, but also investigating the difference in attribution effect made between the model with genetic information and the model without it.

Suppose we have isolates collected from human and non-human cases, of which $H$ isolates belong to humans, and the remaining $N$ isolates are categorized in $J$ groups as the major sources attributed to the infection. Let $I$ genotypes be the total number of unique sequence types (STs) detected from all isolates, and denote $n_j$ as the marginal frequency of STs found in source $j$, where $\sum_j^J n_j = N$. Typically, the number of detected STs $I$ is smaller than the sample size of isolates as multiple isolates will be of the same type.

The likelihood of observing human cases with types $\text{ST}_1, \text{ST}_2, \ldots \text{ST}_H$ follows a multinomial distribution, which via the law of total probability may be expressed as
\begin{align}
  L\Big(\text{ST}_1, \text{ST}_2, \cdots, \text{ST}_H \Big)=\prod_{h=1}^{H}\sum_{j=1}^{J} p\Big(\text{ST}_h \vert \text{source } j\Big) p\Big(\text{source }j\Big),
  \label{Olikh}
\end{align}
where $p(\text{ST}_h \vert \text{source }j)$ is the probability that $\text{ST}_h$ arises from the sampling distribution of source $j$, and $p(\text{source }j)$ is the attribution probability that a random human case is infected from source $j$. Given we know $p(\text{ST}_h \vert \text{source }j)$, estimation of $p(\text{source }j)$ may be found by optimising the likelihood \eqref{Olikh}, for example using a Metropolis Hastings algorithm within a Bayesian context, with suitable priors on $p(\text{source }j)$.

@@

To discover the effect of incorporating genetic information at the allelic profile level as used in the asymmetric Island model, a simple model is developed for the genotype sampling distribution. With the assumption that the observed distribution of genotypes is representative of the true distribution, we model observed genotypes using a multinomial distribution. Let $x_{ij}$ denote the count of ST-$i$ found in source $j$ with the probability $\pi_{ij}$, where $i=1, \ldots, I$ and $j=1, \ldots, J$. To make inference about $\pi_{ij}$, the likelihood is of a multinomial form,
\begin{align}
        L(\boldsymbol{\pi}_{j} ; \boldsymbol{x}_{j}) & = \frac{n_j!}{\prod_{i=1}^{I} x_{ij}!} \prod_{i=1}^I \pi_{ij}^{x_{ij}}, \notag
\end{align}
where $x_{ij}$ can be $0$, indicating STs are yet observed on the sources; $n_j=\sum_{i=1}^I x_{ij}$ is the total count of ST found on source $j$ and $\pi_{ij}$ is subject to $\sum_{i=1}^I \pi_{ij} =1$, and $0 \leq \pi_{ij} \leq 1$. As the family of Dirichlet distributions is a conjugate pair for the multinomial distribution, assume the prior for $\boldsymbol{\pi}_{j}$ follows a Dirichlet density with parameters $\boldsymbol{\alpha}_{j}$,
\begin{align}
  p(\boldsymbol{\pi}_{j}) & \propto \prod_{i=1}^{I}\pi_{ij}^{\alpha_{ij}-1}. \notag
\end{align}
Then the posterior for $\boldsymbol{\pi}_{j}$ takes the form of a Dirichlet probability function with parameters $(\boldsymbol{\alpha}_{j}+\boldsymbol{x}_{j}-\boldsymbol{1})$,
\begin{align}
        p(\boldsymbol{\pi}_{j} \vert \boldsymbol{x}_{j} ) & \propto L(\boldsymbol{\pi}_{j} ; \boldsymbol{x}_{j}) p(\boldsymbol{\pi}_{j}) \notag \\
       & \propto \prod_{i=1}^{I}\pi_{ij}^{\alpha_{ij} + x_{ij}-1}, \quad \alpha_{ij} >0. \notag
 \end{align}
To express the belief that every isolate is equally likely \emph{apriori}, the parameter of the Dirichlet prior is assumed as $\boldsymbol{\alpha}_{j}=\boldsymbol{1}$. Therefore, $p(\text{ST}_h \vert \text{source }j)$, $\boldsymbol{X}$) can be obtained by simulating from the Dirichlet posterior.

Lastly, to estimate the attribution probability, 500 samples are taken from $p(\text{ST}_h \vert \text{source }j)$ generated by the asymmetric Island or simple models. For each posterior sample, we make inferences for $p(\text{source }j)$ using the likelihood~\eqref{Olikh}. This has the effect of integrating over the uncertainty in $p(\text{ST}_h \vert \text{source }j)$ when estimating $p(\text{source }j)$.

@@

The source attribution study described here uses MLST data collected in the Manawatu region of New Zealand from March, 2005 to December, 2014. The sampling was carried out over the same time period and from the same geographical location, providing $H+N=3,480$ isolates taken from human and non-human cases, $H=1,400$ of which are from humans. The non-human isolates are sampled from chicken carcasses, cattle, sheep, environmental water, wild birds, and so on, we therefore put them into four groups representing major sources of infection: poultry, ruminants, water and others. The total number of unique STs typed from all isolates is $I=355$, and 35\% of genotypes are found among human cases. Table~\ref{tab:counts} lists five common genotypes found in human and source isolates. The first four genotypes are frequently observed in human cases. For four sources, ST-45 and ST-474 are commonly detected in poultry, while ST-42 and ST-2026 are typically found in ruminants. Some studies also discover that human cases with ST-45 or ST-474 are likely attributable to poultry, whereas human cases with ST-42 or ST-2026 are more likely attributable to ruminants \cite{Muell, Coll, Cart}. Furthermore, it also shows that there are no observations of ST-2381 in humans, poultry or ruminants, indicating that the genotype may only appear in the water or wild birds.

The data also contain two location variables: one is binary, the other is categorical. The binary variable specifies whether a typed human case is from urban or rural areas, whereas the categorical variable further splits up the areas into seven levels. The levels are highly rural/remote area, rural area with low urban influence, rural area with moderate urban influence, rural area with high urban influence, independent urban area, satellite urban area and main urban area. Table~(\ref{tab2}) lists the number of typed human cases in each classification of rurality. The degree of rurality is a set of integers with a range of [-3,3] corresponding to each defined geographical level. Overall, more than 70\% of typed cases lived in urban areas, and approximately 7.5\% of individuals have no information about the location. We will neglect the missing data in the following analysis by assuming that these cases are randomly missing. Further, the table tells us that the population in 2006 and 2013 for each defined area \cite{Stat1, Stat2} is quite stable. The case rate per 100,000 population is illustrated in Figure~\ref{fig:case_rates}, where we see the burden of infection in urban areas drop remarkably from 2008, coinciding with an intervention in the poultry industry implemented by the New Zealand Food Safety Authority (NZFSA) in 2007 and 2008. It shows the intervention improved infection rates in urban areas, however, it only has a temporary effect in rural areas.

@@

\begin{table}
%  \begin{threeparttable}
  \centering
      \begin{tabular}{rlrrrr}
        \toprule
        Rurality scale & Description & Human cases & 2006 & 2013 \\ \midrule
        -3 & Highly rural/remote area & 19 & 1,572 & 1,527\\
        -2 & Rural area with low urban influence & 92 & 8,373 & 8,304\\
        -1 & Rural area with moderate urban influence & 127 & 10,392 & 10,734\\
        0 & Rural area with high urban influence & 70 & 6,579 & 7,155\\
        1 & Independent urban area & 230 & 28,611 & 28,188\\
        2 & Satellite urban area & 178 & 19,725 & 20,525\\
        3 & Main urban area & 579 & 76,047 & 78,108\\
        \bottomrule
      \end{tabular}
      \caption{Each category of rurality is defined along with the respective number of human cases typed during 2005 and 2014, and the population size in 2006 and 2013, in the Manawatu region of New Zealand.}
      \label{tab2}
\end{table}

\begin{figure}
  \centering
  \includegraphics[width=.8\linewidth]{Figures/case_rates}
  \caption{Case rates per 100,000 population in urban and rural areas of the Manawatu region of New Zealand from 2005 through 2014. An intervention in the poultry industry conducted in 2007 and 2008 resulted in a decreasing incidence of campylobacteriosis in the following years in urban areas.}
  \label{fig:case_rates}
\end{figure}

@@

Previously we described how to estimate the marginal probabilities that a randomly selected human case is due to a given source. To extend this analysis so as to include individual level covariates, we need to calculate subject-specific attribution (conditional) probabilities. To that end, let $F_{jh}$ denote the attribution probability of source $j$ for the $h^\textrm{th}$ human case, with constraints $\sum_{j=1}^J F_j =1$ and $0 \leq F_j \leq 1$, where $j=1, \ldots, J$. We model the probabilities $F_{jh}$ using a linear model on the logit scale, that is,
\begin{align}
  F_{jh}  = \frac{\exp (f_{jh})}{\sum_{j=1}^4 \exp(f_{jh})},
  \label{capF}
\end{align}
where $f_{4h}=0$ is treated as the baseline of $f_{jh}$. Consider the case where the ST data for each human case is supplemented by $p$ additional variables. We develop a general model of $f_{jh}$ with linear combinations of the variables, $c_1, \ldots, c_p$. The model for the $h^\textrm{th}$ individual is then of the form
\begin{align}
f_{jh}  = \alpha_{j} + \beta_{j1} c_{1h} + \beta_{j2} c_{2h} + \ldots + \beta_{jp} c_{ph},  \notag
\end{align}
so that more covariates can be considered in the model to further investigate how the attribution is associated with the factors of interest. Note that if there is a single categorical variable with $L$ levels, then $F_{jh}$ and $f_{jh}$ will take no more than $L$ distinct values. In a slight abuse of notation, we will at times refer to $F_{jh}$ in which the $h$ index refers to the factor level, rather than to a particular subject at that level.

To apply the general model of $f_{jh}$ to the campylobacteriosis data, assume $z$ is the categorical variable representing the classified location of each human case. Then two ways of treating the variable $z$ in model fitting are proposed: one is to treat it as numeric, and the other as categorical. To differentiate the performance between the two fitted models, we link ``the linear model'' and ``the categorical model'' to the first and the latter fitted model, respectively. Hence, the linear prediction function for source $j$ for each human case given the degree of rurality is a numeric variable and can be written as
\begin{align}
  f_{jh} = \alpha_{j} + \beta_{j} z_{h},
  \label{modL}
\end{align}
where $z_{h}$ can be any number of the seven scales if case $h$ was from such a degree of rurality. Conversely, if we treat each of the seven rurality degrees as an indicator with a superscript number $d$, which corresponds with the position of the category ranged from -3 to 3, the model~\eqref{modL} can be rewritten as
\begin{equation}
  f_{jh} = \beta_{1j} z_{1h}+ \beta_{2j} z_{2h} + \ldots + \beta_{dj} z_{dh} + \ldots +  \beta_{7j} z_{7h},
  \label{modC}
\end{equation}
where
\begin{equation*}
z_{dh} =
\left\{\begin{array}{rcl}
1 && \mbox{if case $h$ is in the category $d$,} \notag \\ 0 && \mbox{otherwise}.
\end{array}\right.
\end{equation*}
As a consequence, the estimated attribution probabilities are obtained via equation~\eqref{capF} after fitting the data to model~\eqref{modL} or to model~\eqref{modC}.

@@

\begin{enumerate}
\setcounter{enumi}{-1}
\item Sample $T$ random values from $N(0,1)$ as the initial values points of the parameter set $\boldsymbol{\theta}$ for model~\eqref{modL} or model~\eqref{modC}.
\item Sample a permutation $P_T$ of $\{1, \ldots, T\}$.\label{algorithm_loop}
\item For each $t \in P_T$:
\begin{enumerate}
\item Propose a candidate $\boldsymbol{\theta}^*$ with $\theta_{(t)}$ updated by a normal proposal distribution, $N(\theta_{(t)}, 1)$

\item Use $\boldsymbol{\theta}^*$ to calculate a new set of $f^*$ for source $j$, $j=1, 2, 3$, via model~\eqref{modL} or model~\eqref{modC}, and find the associated $F^*$ for each case by putting the vector $(f^*, f_4=0)$ in equation~\eqref{capF}

>>>>

\begin{document}
\runninghead{Extending statistical models for the source attribution of zoonotic diseases}
\title{Extending statistical models for source attribution of zoonotic diseases: A study of campylobacteriosis}
\author{Sih-Jing Liao\affil{a},
Jonathan Marshall\affil{a}\corrauth\footnotemark[2], Martin L. Hazelton\affil{a} and Nigel P. French\affil{b}}
\address{\affilnum{a} Institute of Fundamental Sciences-Statistics, Massey University, Palmerston North, New Zealand\\
\affilnum{b} mEpiLab, Hopkirk Research Institute, Massey University, Palmerston North, New Zealand}
\corraddr{Jonathan Marshall, Institute of Fundamental Sciences-Statistics, Massey University, Private Bag 11222, Palmerston North, New Zealand\\ \footnotemark[2] E-mail: J.C.marshall@massey.ac.nz}

@@

Preventing and controlling zoonoses through the design and implementation of public health policies requires a thorough understanding of transmission pathways. Modeling jointly the epidemiological data and genetic information of microbial isolates derived from cases provides a methodology for tracing back the source of infection. In this paper, the attribution probability for human cases of campylobacteriosis for each source, conditional on the extent to which each case resides in a rural compared to urban environment, is estimated. A model that incorporates genetic data and evolutionary processes is applied alongside a newly developed genetic-free model. We show that inference from each model is comparable except for rare microbial genotypes, and that the effect of ``rurality'' may be modeled linearly on the logit scale, with increasing rurality leading to increasing likelihood of ruminant-sourced campylobacteriosis.

@@

\section{Introduction}
Modeling of disease surveillance data to explore patterns of infectious diseases has had a long history in public health. Infectious diseases can cause high economic and medical costs due to morbidity and mortality. In recent decades, the annual number of global deaths caused by infections has leveled off at approximate 15 million and may remain at this level for the next three decades \cite{DyeC, WHOM}. In order for such an enormous health burden to be reduced, preventing and controlling infectious diseases becomes extraordinarily important, and our ability to intervene depends on how much we know about the nature of disease transmission. For zoonotic diseases, where transmission to humans from animal reservoirs may be via food, water, through environmental contamination or direct contact with animals, knowledge of the potential sources and pathways of infection is key to reducing the burden of disease. Quantifying the burden of illness attributable to each pathway and source provides valuable information to health professionals making decisions on public health policy for disease control.

Zoonotic transmission from animals to humans may be complex, involving many sources and exposures linked by differing pathways. For instance, infected wild birds may contaminate environmental water and cause disease spread to water users, either humans or other animals \cite{Wagen}. Tracing the source of infection is thus crucial to increasing the ability to implement risk management and intervention \cite{Wilso, Morel}. In terms of statistical analysis, modeling zoonoses requires an advanced approach with the focus changed from just epidemiology to a combination of epidemiology, evolutionary genetics and biology \cite{Muell}. Hence, some source attribution models have been proposed to estimate the number of cases attributable to different sources by utilizing epidemiological information and the association with genotypes found in humans and sources \cite{vanP, Hald, MullA}.

\begin{table}
  \begin{center}
    \begin{tabular}{crrrrrrr}
      \toprule
      ST & \textit{asp}A & \textit{gln}A & \textit{glt}A & \textit{gly}A & \textit{pgm}& \textit{tkt}& \textit{unc}A\\ \midrule
      403 & 10 & 27 & 16 & 19 & 10 & 5 & 7\\
      2026 & 10 & 1 & 16 & 19 & 10 & 5 & 7\\
      474 & 2 & 4 & 1 & 2 & 2 & 1 & 5 \\
      2343 & 2 & 4 & 5 & 2 & 10 & 1 & 5\\
      \bottomrule
    \end{tabular}
  \end{center}
  \caption{The allelic profiles of a selection of genotypes, composed of seven allele numbers at each of the seven loci.}
  \label{tab:mlst}
\end{table}

The genetic information used in such integrated models is typically derived from molecular genotyping that groups closely related organisms together \cite{Cotta}. Typically molecular typing methods, such as multilocus sequence typing (MLST) \cite{Dingl, Coll, Urwin} might be used, which utilizes nucleotide sequences of internal fragments of a small set of housekeeping genes. Such sequences have sufficient variation to distinguish differing pathogen lineages, while being relatively stable within lineages. Each unique nucleotide sequence (allele) at each housekeeping gene (locus) is assigned a number, and the set of numbers across all loci (the allelic profile) is then taken as the genotype, which is assigned a sequence type (ST) number.

@@

In this paper we develop statistical models for source attribution, and demonstrate their use on a case study of human campylobacteriosis conducted in New Zealand \cite{Marsh}. Human campylobacteriosis is caused mainly by \textit{C. jejuni} and \textit{C. coli} which are the dominant species associated with approximately 80\% and 15\% of illnesses, respectively \cite{Guert}. Common symptoms of infection are diarrhea, abdominal pain and fever, however, a severe complication named Guillain Barr\'{e} syndrome may develop, which is a life-threatening disease that weakens the nervous system and leads to paralysis in limbs and respiration \cite{Hanha}. The pathogen can be spread between animals, or from animals and wild birds to humans. Transmission routes may be via drinking contaminated water, eating undercooked animal food products, or handling animal food products that are already contaminated by faeces. In this study, we compare the performance of the asymmetric Island model \cite{Wilso}, which considers genetic evolution when estimating the genotype sampling distribution on each source, to a simple model that utilizes only the prevalence of each type to derive the sampling distribution. We then extend both models in a Bayesian context to incorporate covariates, exploring the effect of human case rurality on attribution results via a linear trend on the logit scale or with separate categories, and performing model comparison.

\section{Models for source attribution with and without microbial genetic information}
\subsection{Notation and assumptions}
The goal of attribution models is to estimate the probability that the observed human cases arise from each putative source. Our data comprise microbial genotype information from each observed human case obtained from analysis of stool samples, and also from a pool of non-human cases corresponding to potential zoonotic sources of disease

Given this genotyping information, we first estimate the sampling distribution of genotypes for each source, and then estimate the appropriate combinations of those genotype distributions that most likely give rise to the set of genotypes observed among human cases. Specifying first the sampling distribution of genotypes found on sources is fundamental for the purpose of not only exploring how it affects the source attribution probability, but also investigating the difference in attribution effect made between different genetic models.

Suppose we have isolates collected from human and non-human cases, of which $H$ isolates belong to humans, and the remaining $N$ isolates are categorized in $J$ groups as the major sources attributed to the infection. Let $I$ genotypes be the total number of unique types detected from all isolates, and denote $n_j$ as the marginal frequency of types found in source $j$, where $\sum_j^J n_j = N$. Typically, the number of detected types $I$ is smaller than the sample size of isolates as multiple isolates will be of the same type.

The likelihood of observing human cases with types $\text{ST}_1, \text{ST}_2, \ldots \text{ST}_H$ follows a multinomial distribution, which via the law of total probability may be expressed as
\begin{align}
  L\Big(\text{ST}_1, \text{ST}_2, \cdots, \text{ST}_H \Big)=\prod_{h=1}^{H}\sum_{j=1}^{J} p\Big(\text{ST}_h \vert \text{source } j\Big) p\Big(\text{source }j\Big),
  \label{Olikh}
\end{align}
where $p(\text{ST}_h \vert \text{source }j)$ is the probability that genotype $\text{ST}_h$ arises from the sampling distribution of source $j$, and $p(\text{source }j)$ is the attribution probability that a random human case is infected from source $j$. Given we know $p(\text{ST}_h \vert \text{source }j)$, estimation of $p(\text{source }j)$ may be found by optimising the likelihood \eqref{Olikh}, for example using a Metropolis Hastings algorithm within a Bayesian context, with suitable priors on $p(\text{source }j)$.

@@

To discover the effect of incorporating genetic information at the allelic profile level as used in the asymmetric Island model, a simple model is developed for the genotype sampling distribution. With the assumption that the observed distribution of genotypes is representative of the true distribution, we model observed genotypes using a multinomial distribution. Let $x_{ij}$ denote the count of genotype ST-$i$ found in source $j$ with the probability $\pi_{ij}$, where $i=1, \ldots, I$ and $j=1, \ldots, J$. To make inference about $\pi_{ij}$, the likelihood is of a multinomial form,
\begin{align}
        L(\boldsymbol{\pi}_{j} ; \boldsymbol{x}_{j}) & = \frac{n_j!}{\prod_{i=1}^{I} x_{ij}!} \prod_{i=1}^I \pi_{ij}^{x_{ij}}, \notag
\end{align}
where $x_{ij}$ can be $0$, indicating genotypes are yet observed on the sources; $n_j=\sum_{i=1}^I x_{ij}$ is the total count of all types found on source $j$ and $\pi_{ij}$ is subject to $\sum_{i=1}^I \pi_{ij} =1$, and $0 \leq \pi_{ij} \leq 1$. As the family of Dirichlet distributions is a conjugate pair for the multinomial distribution, assume the prior for $\boldsymbol{\pi}_{j}$ follows a Dirichlet density with parameters $\boldsymbol{\alpha}_{j}$,
\begin{align}
  p(\boldsymbol{\pi}_{j}) & \propto \prod_{i=1}^{I}\pi_{ij}^{\alpha_{ij}-1}. \notag
\end{align}
Then the posterior for $\boldsymbol{\pi}_{j}$ takes the form of a Dirichlet probability function with parameters $(\boldsymbol{\alpha}_{j}+\boldsymbol{x}_{j}-\boldsymbol{1})$,
\begin{align}
        p(\boldsymbol{\pi}_{j} \vert \boldsymbol{x}_{j} ) & \propto L(\boldsymbol{\pi}_{j} ; \boldsymbol{x}_{j}) p(\boldsymbol{\pi}_{j}) \notag \\
       & \propto \prod_{i=1}^{I}\pi_{ij}^{\alpha_{ij} + x_{ij}-1}, \quad \alpha_{ij} >0. \notag
 \end{align}
To express the belief that every isolate is equally likely \emph{apriori}, the parameter of the Dirichlet prior is assumed as $\boldsymbol{\alpha}_{j}=\boldsymbol{1}$. Therefore, $p(\text{ST}_h \vert \text{source }j)$, $\boldsymbol{X}$) can be obtained by simulating from the Dirichlet posterior.

Lastly, to estimate the attribution probability, 500 samples are taken from $p(\text{ST}_h \vert \text{source }j)$ generated by the asymmetric Island or simple models. For each posterior sample, we infer $p(\text{source }j)$ using the likelihood~\eqref{Olikh}. This has the effect of integrating over the uncertainty in $p(\text{ST}_h \vert \text{source }j)$ when estimating $p(\text{source }j)$.

@@

The source attribution study described here uses human case data ($H=1,400$) collected in the Manawatu region of New Zealand from March 2005 to December 2014. In addition, over the same time period and from the same geographical location, isolates $N=2,080$ sampled from chicken carcasses, cattle, sheep, environmental water, wild birds, and so on, were also taken, and categorised into four groups  representing major sources of infection: poultry, ruminants, water and others. The total number of unique genotypes from all isolates is $I=355$, with 35\% of genotypes found among human cases. Table~\ref{tab:counts} lists five common genotypes found in human and source isolates, the first four of which are frequently observed in human cases. As found in other studies, ST-45 and ST-474 are detected mainly in poultry, while ST-42 and ST-2026 are detected mainly in ruminants\cite{Muell, Coll, Cart}. The fifth genotype, ST-2381 is not found among human cases, appearing only in the water and other sources, in this case being found in Pukeko and Takah\={e} birds from the Rallidae family \cite{french2014evolution}.

The data also contains location information, in the form of an ordinal classification of urban and rural areas with seven levels coded from -3 to 3: highly rural/remote area, rural area with low urban influence, rural area with moderate urban influence, rural area with high urban influence, independent urban area, satellite urban area and main urban area. Approximately 7.5\% of individuals in the Manawatu dataset contain no information about the location, which we assume are missing at random. Table~\ref{tab2} lists the remainder of typed human cases in each classification of rurality as well as the population from the 2006 and 2013 Census\cite{Stat1, Stat2}. The case rate per 100,000 population is illustrated in Figure~\ref{fig:case_rates}, where we see the burden of infection in urban areas drop remarkably from 2008, coinciding with an intervention in the poultry industry implemented by the New Zealand Food Safety Authority (NZFSA) in 2007 and 2008. It shows the intervention improved infection rates in urban areas, however, it only has a temporary effect in rural areas.

@@

\begin{table}
%  \begin{threeparttable}
  \centering
      \begin{tabular}{rlrrrr}
        \toprule
        Rurality scale & Description & Human cases & 2006 & 2013 \\ \midrule
        -3 & Highly rural/remote area & 19 & 1,572 & 1,527\\
        -2 & Rural area with low urban influence & 92 & 8,373 & 8,304\\
        -1 & Rural area with moderate urban influence & 127 & 10,392 & 10,734\\
        0 & Rural area with high urban influence & 70 & 6,579 & 7,155\\
        1 & Independent urban area & 230 & 28,611 & 28,188\\
        2 & Satellite urban area & 178 & 19,725 & 20,525\\
        3 & Main urban area & 579 & 76,047 & 78,108\\
        \bottomrule
      \end{tabular}
      \caption{The number of human cases in each rurality class from 2005--2014, and the population size in 2006 and 2013, in the Manawatu region of New Zealand.}
      \label{tab2}
\end{table}

\begin{figure}
  \centering
  \includegraphics[width=.8\linewidth]{Figures/case_rates}
  \caption{Case rates per 100,000 population in urban and rural areas of the Manawatu region of New Zealand from 2005 through 2014. An intervention in the poultry industry conducted in 2007 and 2008 resulted in a decreasing incidence of campylobacteriosis in the following years, particularly in urban areas.}
  \label{fig:case_rates}
\end{figure}

@@

Previously we described how to estimate the marginal probabilities that a randomly selected human case is due to a given source. To extend this analysis so as to include individual level covariates, we need to calculate subject-specific attribution (conditional) probabilities. To that end, let $F_{jh}$ denote the attribution probability of source $j$ for the $h^\textrm{th}$ human case, with constraints $\sum_{j=1}^J F_{jh} =1$ and $0 \leq F_{jh} \leq 1$, where $j=1, \ldots, J$. We model the probabilities $F_{jh}$ using a linear model on the logit scale, that is,
\begin{align}
  F_{jh}  = \frac{\exp (f_{jh})}{\sum_{j=1}^4 \exp(f_{jh})},
  \label{capF}
\end{align}
where $f_{4h}=0$ is treated as the baseline of $f_{jh}$. Consider the case where the genotype data for each human case is supplemented by $p$ additional variables. A general model of $f_{jh}$ with linear combinations of the variables, $c_1, \ldots, c_p$, then has the form
\begin{align}
f_{jh}  = \alpha_{j} + \beta_{j1} c_{1h} + \beta_{j2} c_{2h} + \ldots + \beta_{jp} c_{ph},  \notag
\end{align}
for the $h$-th individual. Note that if there is a single categorical variable with $L$ levels, then $F_{jh}$ and $f_{jh}$ will take no more than $L$ distinct values. In a slight abuse of notation, we will at times refer to $F_{jh}$ in which the $h$ index refers to the factor level, rather than to a particular subject at that level.

To apply the general model of $f_{jh}$ to the campylobacteriosis data, assume $z$ is the variable ranging from -3 to 3 representing the classified location of each human case. Then two ways of treating the variable $z$ in model fitting are proposed: one is to treat it as numeric, and the other as categorical. To differentiate the performance between the two fitted models, we link ``the linear model'' and ``the categorical model'' to the first and the latter fitted model, respectively. Hence, the linear prediction function for source $j$ for each human case given the degree of rurality is a numeric variable and can be written as
\begin{align}
  f_{jh} = \alpha_{j} + \beta_{j} z_{h},
  \label{modL}
\end{align}
where $z_{h}$ can be any number of the seven scales if case $h$ was from such a degree of rurality. Conversely, if we treat each of the seven rurality degrees as an indicator with a superscript number $d$, which corresponds with the position of the category ranged from -3 to 3, the model~\eqref{modL} can be rewritten as
\begin{equation}
  f_{jh} = \beta_{1j} z_{1h}+ \beta_{2j} z_{2h} + \ldots + \beta_{dj} z_{dh} + \ldots +  \beta_{7j} z_{7h},
  \label{modC}
\end{equation}
where
\begin{equation*}
z_{dh} =
\left\{\begin{array}{rcl}
1 && \mbox{if case $h$ is in the category $d$,} \notag \\ 0 && \mbox{otherwise}.
\end{array}\right.
\end{equation*}
As a consequence, the estimated attribution probabilities are obtained via equation~\eqref{capF} after fitting the data to model~\eqref{modL} or to model~\eqref{modC}.

@@

\begin{enumerate}
\setcounter{enumi}{-1}
\item Sample $T$ random values from $N(0,1)$ as initial values of the parameter set $\boldsymbol{\theta}$ for model~\eqref{modL} or model~\eqref{modC}.
\item Sample a permutation $P_T$ of $\{1, \ldots, T\}$.\label{algorithm_loop}
\item For each $t \in P_T$:
\begin{enumerate}
\item Propose a candidate $\boldsymbol{\theta}^*$ with $\theta_{(t)}$ updated by a normal proposal distribution, $Q(\theta_{(t)}^*, \theta_{(t)}) = N(\theta_{(t)}, 1)$.

\item Use $\boldsymbol{\theta}^*$ to calculate a new set of $f^*$ for source $j$, $j=1, 2, 3$, via model~\eqref{modL} or model~\eqref{modC}, and find the associated $F^*$ for each case by putting the vector $(f^*, f_4=0)$ in equation~\eqref{capF}.
